---
title: "One_Way_MoPlot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One_Way_MoPlot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MoPlot)
```
# MoPlotOneWay

`MoPlotOneWay()` is a function for visualizing the results of a linear model with one categorical predictor and one numeric dependent variable. It takes as input a model created with `lm()` and produces a plot that highlights group means with their associated uncertainty using bars, and visually emphasizes significant contrasts through colored lines. Optionally, the raw data points can be displayed in the plot. A secondary panel shows the model coefficients, including standardized beta values and effect sizes such as Cohen’s *d*, allowing for a clear and informative summary of the model's findings.  `MoPlotOneWay()` function supports a variety of contrast coding schemes for categorical predictors, including treatment coding (dummy coding), sum to zero coding, Helmert coding, reverse Helmert coding, polynomial coding, sliding difference coding, and customized coding.

### 1 Default Treatment coding

Treatment coding, or dummy coding is a common way to represent categorical variables in regression models. It transforms a categorical predictor with k levels into k − 1 binary (0/1) variables, each comparing one level to a reference (baseline) category.
The following example uses the classic iris dataset with treatment contrasts and shows the plot without raw data points or coefficient panel, using a significance level of 0.05:

```{r, fig.height=7, fig.width=12} 
library(MoPlot)

# Using Iris data
data(iris)
iris$Species <- factor(iris$Species)
contrasts(iris$Species) <- contr.treatment(3)

# Fit the model
model <- lm(Sepal.Length ~ Species, data = iris)

# MoPlotting
MoPlotOneWay(model, data = FALSE, coef = FALSE, siglevel = 0.05)
```

Here is the same plot, but including raw data points and the coefficient panel, also with a 0.05 significance level:
```{r, fig.height=7, fig.width=12, warning=FALSE} 
# MoPlotting
MoPlotOneWay(model, data = TRUE, coef = TRUE, siglevel = 0.05)
```

### 2 Sum coding

Sum coding is another way to represent categorical variables in regression models. Unlike treatment coding, sum coding compares each level to the overall mean rather than a specific reference group. The following example uses again the iris dataset with sum contrasts, showing the plot without raw data points or coefficient panel at a significance level of 0.1

```{r, fig.height=7, fig.width=12} 
# Using Sum coding
contrasts(iris$Species) <- contr.sum(3)

# Fit the model
model <- lm(Sepal.Length ~ Species, data = iris)

# MoPlotting
MoPlotOneWay(model, data = FALSE, coef = FALSE, siglevel = 0.1)
```

Here is the same plot including raw data points and the coefficient panel:
```{r, fig.height=7, fig.width=12, warning=FALSE} 
# MoPlotting
MoPlotOneWay(model, data = TRUE, coef = TRUE, siglevel = 0.1)
```

### 3 Helmert coding

Helmert coding is a contrast coding scheme where each level of a categorical variable is compared to the mean of previous levels. The following example uses a generated dataset (using `DatasetOne()`) with Helmert contrasts. It first shows a plot without raw data points or coefficient panel at a significance level of 0.05:

```{r, fig.height=7, fig.width=12} 
# Using a custom function to generate data with four groups
DT <- DatasetOne(n_sample = 20, means = c(120, 135, 115, 120), sds = 20)
DT$Ind <- as.factor(DT$Ind)
contrasts(DT$Ind) <- contr.helmert(4)

# Fit the model
model <- lm(Dep ~ Ind, data = DT)

# MoPlotting sum
MoPlotOneWay(model, data = FALSE, coef = FALSE, siglevel = 0.05)
```

### 4 Reverse Helmert coding
Reverse Helmert coding compares each level to the mean of the subsequent levels. It is the reverse of Helmert coding and can be useful in some ordered categorical predictors. The example below uses the same custom dataset as before with reverse Helmert contrasts, showing the plot without raw data or coefficient panel at a 0.05 significance level:

```{r, fig.height=7, fig.width=12, warning=FALSE} 
# Using Reverse Helmert coding
ReverseHelmert <- function(n) {
  mat <- matrix(0, n, n - 1)
  for (i in 2:n) {
    mat[i, i - 1] <- 1
    mat[1:(i - 1), i - 1] <- -1 / (i - 1)
  }
  return(mat)
}
contrasts(DT$Ind) <- ReverseHelmert(4)

# Fit the model
model <- lm(Dep ~ Ind, data = DT)

# MoPlotting reverse Helmert
MoPlotOneWay(model, data = FALSE, coef = FALSE, siglevel = 0.05)
```
Here is the same plot including raw data points and the coefficient panel:
```{r, fig.height=7, fig.width=12, warning=FALSE} 
MoPlotOneWay(model, data = TRUE, coef = TRUE, siglevel = 0.05)
```

### 5 Polynomial coding
Polynomial coding tests for trends across ordered factor levels. It compares levels based on polynomial contrasts (linear, quadratic, cubic, etc.). The following example uses polynomial contrasts on the custom dataset, showing the plot without raw data or coefficients at significance level 0.05:

```{r, fig.height=7, fig.width=12, warning=FALSE} 
# Using Polynomial coding
contrasts(DT$Ind) <- contr.poly(4)

# Fit the model
model <- lm(Dep ~ Ind, data = DT)

# MoPlotting polynomial contrasts
MoPlotOneWay(model, data = FALSE, coef = FALSE, siglevel = 0.05)
```

And including raw data and coefficients:
```{r, fig.height=7, fig.width=12, warning=FALSE} 
MoPlotOneWay(model, data = TRUE, coef = TRUE, siglevel = 0.05)
```

### 6 Sliding Difference coding
Sliding difference coding compares adjacent levels of the categorical variable, useful to detect stepwise differences. The following example applies sliding difference contrasts:

```{r, fig.height=7, fig.width=12, warning=FALSE} 
# Using Sliding Difference coding
contrasts(DT$Ind) <- MASS::contr.sdif(4)

# Fit the model
model <- lm(Dep ~ Ind, data = DT)

# MoPlotting sliding difference contrasts
MoPlotOneWay(model, data = FALSE, coef = FALSE, siglevel = 0.05)
```

With raw data and coefficients included:
```{r, fig.height=7, fig.width=12, warning=FALSE} 
MoPlotOneWay(model, data = TRUE, coef = TRUE, siglevel = 0.05)
```

### 7 Customized coding
MoPlotOneWay() also supports customized contrast coding schemes. You can specify your own contrast matrix for a factor variable. Below is an example of using custom contrasts:

```{r, fig.height=7, fig.width=12, warning=FALSE} 
# Define custom contrasts
custom_contrasts <- matrix(c(
  1,  1,  0,
  1,  -1,  0,
  -1, 0,  1,
  -1,  0, -1
), ncol = 3, byrow = TRUE)
rownames(custom_contrasts) <- levels(DT$Ind)
colnames(custom_contrasts) <- c("C1", "C2", "C3")
contrasts(DT$Ind) <- custom_contrasts

# Fit the model
model <- lm(Dep ~ Ind, data = DT)

# MoPlotting with customized contrasts
MoPlotOneWay(model, data = FALSE, coef = FALSE, siglevel = 0.05)
```

With raw data and coefficient panel:
```{r, fig.height=7, fig.width=12, warning=FALSE} 
MoPlotOneWay(model, data = TRUE, coef = TRUE, siglevel = 0.05)
MoPlotOneWay(model, data = TRUE, coef = TRUE, siglevel = 0.05, facet.pers = F)

```


Without faceting:

```{r, fig.height=7, fig.width=12, warning=FALSE} 
MoPlotOneWay(model, data = TRUE, coef = TRUE, siglevel = 0.05, facet.pers = F)
```
